Learning to Optimize Real-World Programs

### 2021-02-05 paper.tex
update

%\usepackage{lstlinebgrd}
\usepackage{xspace}
\usepackage{soul}
\usepackage{color}

@@

@@

%\pcyin{DISCUSSED IN SLACK: Can we sell our paper as ``learning to optimize real-world \textbf{C} programs''? It sounds fancier. I understand what we are actually doing is to learn to optimize X86-64 assembly. But since we mention in \cref{sec:benchmark} that we are optimizing C programs. It might sound confusing for ML people who do not know much about program optimization:joycryface:.}
\pcyin{Throughout the paper I mistakenly used the term ``Learning to optimize C programs''. I will revise accordingly.}

@@

Because each program is a function that maps inputs to outputs, we will use $I$ to represent the hardware state prior to executing the program (i.e., input) and $O$ to represent the hardware state after executing the program (i.e., output).
Specifically, our goal is to learn a model $f_{\theta}: \text{F}_u \mapsto \text{F}_o$ such that a model-generated (optimized) program $\hat{F_o}$ attains lower cost under a cost function $\mathcal{C}(\cdot)$ evaluated on a suite of $K$ input-output test cases $\{ IO \}_{k=1}^K$:
%on our synthesized program $\hat{\text{F}}_o$,
%and observing $\mathcal{C}$ which may measure approximate runtime, energy consumption, or other goals. Furthermore, we may evaluate our distance function $\mathcal{D}$ either as equivalence under all test cases, or through verification by using an SMT solver. We consider a program optimized when: 
% cost function $\mathbfcal{C}$ which is able to evaluate desired program behavior, such as approximate runtime or energy consumption. For the transductive objective, the learned model's objective is then the following: 
\begin{equation}
    \label{eqn:optimizaiton_goal}
    \begin{split}
        \mathbfcal{C} \Big(\hat{\textrm{F}}_{o}; \{IO_k\}_{k=1...K} \Big)  \ 
        < \ 
        \mathbfcal{C} \Big(\textrm{F}_{u}; \{IO_k\}_{k=1...K}\Big) \\
         s.t. \;\; \mathbfcal{D}( \ 
                        \hat{\textrm{F}}_{o}, \textrm{F}_{u} ) \ 
                        = 0  % \;\; \
        %  \hat{\textrm{F}}_{opt}^i(I_k^i) = \ 
        %         O_k^i , \\
            % \forall  \; k \;  \in 1...K 
    \end{split}
\end{equation}

where $\mathcal{D}(\cdot) \in \{ 0, 1 \}$ is an indicator function that measures if the model optimized program $\hat{F}_o$ is functionally equivalent to the original one $\p$. 
An intuitive choice of $\mathcal{D}$ is directly executing $\p$ and $\hat{F}_o$ on the test cases $\{ IO \}_{k=1}^K$ and checking outputs equivalence. The cost function $\mathcal{C}(\cdot)$ considers metrics like runtime and the number of instructions in a program.
We will present more details in \cref{sec:setup}.
%, and could be computed by directly executing the optimized program on the test cases $\{ IO \}_{k=1}^K$. 
%Additionally, we explain how $\mathcal{D}$ may be calculated by using the suite of test cases, or by incorporting other methods like SMT-based solvers. 
For conciseness, we will abbreviate $\mathbfcal{C} (\hat{\textrm{F}}_{o}; \{IO_k\}_{k=1}^K)$ as $\mathbfcal{C} (\hat{\textrm{F}}_{o})$.

@@

%We perform a two-step process for learning program optimization.
In this section we elaborate on our approach that learns to optimize real-world C programs. 
we first describe the underlying neural model before introducing the learning algorithm.

@@

Our program optimization model $f_{\theta}$ is a neural sequence-to-sequence network, where input (original) and output (optimized) programs are represented as sequences of tokens.
Specifically, $f_{\theta}$ is parameterized with a standard Transformer-based encoder \citep{vaswani2017attention}, but with a LSTM-based decoder \citep{hochreiter1997long,chen-etal-2018-best} instead.
This was done to speed-up prediction within the iterative training loop described in \S\ref{sec:hillclimbing}, as Transformers suffer from quadratic decoding time in sequence length \citep{zhang-etal-2018-accelerating}. We defer more implementation-specific details for our experiments to Section \ref{sec:experiments}.

@@

\subsection{Learning Algorithm}

@@

%We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
%It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
\cref{alg:hill_climbing} illustrates the hill-climbing algorithm for learning to optimize programs.
It consists of two steps, an exploration step (lines xxx-xxx) where the model seeks to discover alternative optimizations that are more efficient than the compiler generated targets used in pre-training, and a learning step (lines xxx-xxx), where the model parameters are updated using newly discovered optimized programs\pcyin{TODO: mark those two steps in the Algorithm}.
First, in the exploration step, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and execute $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
%Each output from the batch is executed on the test cases to test equivalence and its performance characteristics are measured and profiled. 
If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
\pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
After the exploration step is taken, in the learning step, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

\paragraph{Discussion} \cref{alg:hill_climbing} bears some resemblance to self-imitation learning \cite{oh2018self} \pcyin{can you briefly explain why they are similar?}; except we train on the entire sequence, instead of individual actions within a trajectory. Learning with this algorithm may be interpreted as imitation learning on an expert policy that iteratively improves. 
More formally, it could be interpreted as an extreme case of normalized off-policy reinforcement learning. We refer interested readers to Apendix xxx for more details \alex{TODO: write out and put in bibliography.}. 
Finally, we remark that an alternative approach for learning program optimizations could be policy gradient methods.
In this paper we adopt a simpler hill-climbing approach as we observe in our pilot experiments that policy gradient is significantly more unstable, due to the general sparsity of rewards and the long trajectories of programs (\pyin{TODO: xxx} tokens in average).
Such instability of RL-based learning algorithms confirms with existing research on program synthesis \citep{bunel2018leveraging}.

@@

\textcolor{red}{=== Begin of Alex's Version ===}

%\pcyin{When explaining a model/algorithm, it's important to first briefly describe the intuition behind each component/sub-section at the very beginning. Can we say something like the follows here: With the model pre-trained on compiler-optimized examples, we then aim to learn more efficient optimization strategies by directly minimizing the expected cost of model-optimized programs.}

%%Commented by Pengcheng: 
Policy gradient algorithms seem like a natural fit for learning how to optimize programs; however, methods based on maximum-likelihood estimation, or imitation learning, provide advantages in terms of stability. Training large neural networks with policy gradient algorithms can be difficult to tune. These difficulties are often attributed to the gradient's high variance which is exacerbated by other issues in program synthesis: a general sparsity of rewards, credit assignment over long sequences, and brittleness of program semantics.

% \alex{I think this point here contradicts my earlier point about needing to sample from a grammar.... If a grammar really isn't that necessary after pre-training, then is brittle syntax really an issue?? It may be good to leave this out or accept either the nuance proposed / contradiction}
% \pcyin{Perhaps just removing ``brittleness of program syntax'' is good?}

%Commented by Pengcheng: 
If an expert policy or expert examples are available, imitation learning can be desirable, because of its stability. Impressive results have been achieved in the game of computer Go and other games by minimizing the cross-entropy between a policy and an expert-like Monte Carlo tree search policy \cite{silver2017mastering}. In program synthesis from I/O example tasks, supervised learning on example programs has been shown to be competitive, and at times even superior to RL \cite{bunel2018leveraging}. The effectiveness of supervised, or imitation learning, on program synthesis tasks is even more surprising, given the programs used as targets were generated from random generation from a grammar. Given that real-world programs demonstrate a degree of naturalness, and thus, a higher degree of predictability and repetition \cite{hindle2012naturalness}, one may expect imitation learning to generally perform even better in practice on general program synthesis tasks. \alex{I feel these 2 sentences are good points but could be a little too much detail.}\pcyin{I agree. We could try shortening this paragraph once Section 6 is nearly done.}

% The experiments performed in \cite{bunel2018leveraging} showed that in program synthesis tasks in the Karel domain, maximum-likelihood training on just one correct program example was competitive with reinforcement learning on a reward function defined by executed program outputs. Moreover, maximum-likelihood training outperformed reinforcement learning when the beam size for sampling candidates was large. The result may be surprising as maximum-likelihood training ignores issue of  \textit{program aliasing} in which multiple correct programs, not just one, is capable of satisfying the program specification, and may over-fit on the maximum-likelihood target.

% Commented by Pengcheng

Inspired by the success of imitation learning in program synthesis, in this paper we propose an intuitive hill-climbing algorithm, a special case of imitation learning, for learning program optimizations. 
%We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
%It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
It consists of a two-step process. 
First, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and run $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
%Each output from the batch is executed on the test cases to test equivalence and its performance characteristics are measured and profiled. 
If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
\pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
After the exploration step is taken, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

Algorithm \ref{alg:hill_climbing} bears some resemblance to self-imitation learning \cite{oh2018self} \pcyin{can you briefly explain why they are similar?}; except we train on the entire sequence, instead of individual actions within a trajectory. Learning with this algorithm may be interpreted as imitation learning on an expert policy that iteratively improves. In our appendix, we also show how this is an extreme case of normalized off-policy reinforcement learning \alex{TODO: write out and put in bibliography.}.

\textcolor{red}{=== End of Alex's Version ===}

\begin{algorithm}[t]
\begin{algorithmic}
\STATE Initialize model $\textbf{Model}$ parameters $\theta$ from pre-trained model \\
\STATE Initialize dataset of program function pairs and test cases: \\  
    $D = D_o =  \Big\{ \big( \
    \{IO_k^i\}_{k=1...K}, \
    \textrm{F}_{u}^i, \
    \textrm{F}_{o}^i, \
    % \textit{LiveOut}^i\
    \big) \ 
    \Big\}_{i=1...N}$ \\
 \WHILE{not converged}
  \STATE Sample a batch $B_{\mathrm{ex}}$ from $D_o$\\
  \FOR{ $\big( \
    \{IO_k^i\}_{k=1...K}, \
    \textrm{F}_{u}^i, \
    \textrm{F}_{o}^i, \
    % \textit{LiveOut}^i\
    \big)$  in $B_{\mathrm{ex}}$} 
    \STATE  sample $\hat{\textrm{F}}_{o}^i \; \sim \; f_{\theta}(\textrm{F}_{u}^i)$ \\
    \STATE calculate 
    $\mathbfcal{C}(\textrm{F}_{o}^i)$, 
    and $\mathbfcal{D} (\hat{\textrm{F}}_{o}^i, \textrm{F}_{o}^i)$ 
    \IF{$\mathbfcal{D} (\hat{\textrm{F}}_{o}^i, \textrm{F}_{o}^i) == 0$ and \ 
    $\mathbfcal{C}(\hat{\textrm{F}}_{o}^i) < \ 
    \mathbfcal{C}(\textrm{F}_{o}^i) $ }
        \STATE  Replace $\textrm{F}_{o}^i$ with sample $\hat{\textrm{F}}_{o}^i$ in $D$ \\
    \ENDIF
  \ENDFOR
  \STATE Sample a batch $B_{\mathrm{tr}}$ from $D$\\
  \STATE Update $\theta$ via supervised learning on $B_{\mathrm{tr}}$ from $D$ \\
 \ENDWHILE
 \caption{Hill Climbing for Program Optimization}
 \label{alg:hill_climbing}
\end{algorithmic}
\end{algorithm}

@@

\paragraph{Actor-Learner Architecture}
%\gn{Is this an architecture or algorithm?}
\gn{I changed this to a paragraph, as it's part of the iterative hill-climbing approach.}

\pcyin{Maybe briefly explain why we use multiprocessing: because evaluating $\mathcal{D}(\cdot)$ is slow. Would be better to give a rough runtime number.}
Importantly, to perform our experiments at scale, we utilize an actor-learner set up  \citep{liang2018memory, espeholt2018impala}. Specifically, before the training process begins, multiple actor threads are created which inherit the parameters of the parent learner and the dataset. For every iteration, each of the actor threads independently samples $B_{\textrm{ex}}$ from the distribution of the inherited model.

@@

Utilizing the actor-learner architecture was pivotal in being able to scale the speed of training. Otherwise, during our experiments, performing synchronized learning was slow \alex{is there a good way to quantify this ?}.

@@

We are primarily interested in answering the following research questions:

(1) Are neural sequence-to-sequence models capable of learning generalized semantics-preserving program optimizations ? We seek to measure this by investigating the model's ability to transduce correct programs that outperform unoptimized programs

(2) Can neural sequence model pre-trained on demonstrations of an optimizing compiler like \texttt{GCC} (\cref{sec:sec:results_training_with_test_cases}) learn to outperform the optimization levels for which it was pre-trained ?

(3) In learning to optimize real world programs, how much benefit do more involved methods of program verification, such as using SMT solvers, provide ?

(4) What kind of optimization patterns does the neural optimizer learn (\cref{sec:sec:opt_examples})?

@@

\paragraph{Measuring Program Performance and Correctness} 
%\pcyin{We need to keep in mind that since we chose not to report results from training with SMT-solvers, we need to be careful to distinguish between the correctness function $\mathcal{D}$ used in \cref{alg:hill_climbing} for training and the SMT-bsed post-verification used in Table 2.}
In our experiments, we calculate the cost function $\mathcal{C}$ (E.q. \ref{eqn:optimizaiton_goal}) first by calculating the sum of all expected CPU clock-cycles for each opcode in the synthesized assembly. 
Then, we execute the entire test suite on the synthesized program and count the number of instructions actually executed.%
\footnote{We chose to use these these methods of approximating runtime, as the actual measured runtime of the program re-writes may both be hardware-dependent and inconsistent between executions due to idiosyncrasies of the hardware used for experiments.}
For each unoptimized program $F_u$, have a corresponding set of input examples ${\{IO_k\}}_{k=1...K}$ generated through random generation. \pcyin{The last sentence is a bit confusing?}

@@

\paragraph{Evaluation}
Given an evaluation set $E$ of unoptimized programs, we report the percentage of programs in $E$ whose model-optimized programs outperform the compiler-optimized results as measured by the cost function $C(\cdot)$.
Specifically, for each unoptimized program, we generate model-optimized candidates using a beam-size of 10, executing all on the test suite $\{IO\}_{k}^K$ to evaluate the cost $\mathcal{C}$ and correctness, and report results on the best-performing correct sample (if one was available). 
%We compare the model optimized program with two compiler configurations.

@@

% \subsection{Training Only with test cases}
\subsection{Training using $\bm{\mathcal{D}_\textrm{I/O}}$}
\label{sec:sec:results_training_with_test_cases}

@@

In trying to learn program optimizations, we first train the model using the hill-climbing algorithm (\cref{alg:hill_climbing}) where we use test cases to evaluate program correctness (\correctfuncio/). Table \ref{tab:random_generation_results} lists the results.

We first find that our model is able to find new optimizations for this task, wherein on the 24.27\% of programs sampled for the transductive task were able to outperform the \texttt{GCC -Og} benchmark. This performance on the \texttt{GCC -Og} benchmark also seemed to generalize to the hold-out test set; and the performance on the \texttt{GCC -O0} task also seemed to generalize as well. An interesting result was that the hill-climbing algorithm performed slightly worse in the inductive task of outperforming the \texttt{GCC -O0} unoptimized baseline.

>>>>

%\usepackage{lstlinebgrd}
\usepackage{xspace}
\usepackage{soul}
\usepackage{color}
\usepackage{multirow}

@@

\Crefname{algorithm}{Algo.}{Algorithms}
\Crefname{table}{Tab.}{Tables}
\crefname{section}{\S}{\S\S}
\crefname{subsection}{\S}{\S\S}

\sidecaptionvpos{figure}{t}

\usetikzlibrary{tikzmark}

\renewcommand{\sectionautorefname}{\S}
\renewcommand{\subsectionautorefname}{\S}
\renewcommand{\subsubsectionautorefname}{\S}
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\tableautorefname}{Tab.}

@@

%\pcyin{DISCUSSED IN SLACK: Can we sell our paper as ``learning to optimize real-world \textbf{C} programs''? It sounds fancier. I understand what we are actually doing is to learn to optimize X86-64 assembly. But since we mention in \cref{sec:benchmark} that we are optimizing C programs. It might sound confusing for ML people who do not know much about program optimization:joycryface:.}
% \pcyin{Throughout the paper I mistakenly used the term ``Learning to optimize C programs''. I will revise accordingly.}

@@

Because each program is a function that maps inputs to outputs, we will use $I$ to represent the hardware state prior to executing the program (i.e., input) and $O$ to represent the hardware state after executing the program (i.e., output).
Specifically, our goal is to learn a model $f_{\theta}: \text{F}_u \mapsto \text{F}_o$ such that a model-generated (optimized) program $\hat{F_o}$ attains lower cost under a cost function $\mathcal{C}(\cdot)$ evaluated on a suite of $K$ input-output test cases $\{ IO \}_{k=1}^K$:
%on our synthesized program $\hat{\text{F}}_o$,
%and observing $\mathcal{C}$ which may measure approximate runtime, energy consumption, or other goals. Furthermore, we may evaluate our distance function $\mathcal{D}$ either as equivalence under all test cases, or through verification by using an SMT solver. We consider a program optimized when: 
% cost function $\mathbfcal{C}$ which is able to evaluate desired program behavior, such as approximate runtime or energy consumption. For the transductive objective, the learned model's objective is then the following: 
\begin{equation}
    \label{eqn:optimizaiton_goal}
    \begin{split}
        \mathcal{C} \Big(\hat{\textrm{F}}_{o}; \{IO_k\}_{k=1}^k \Big)  \ 
        < \ 
        \mathcal{C} \Big(\textrm{F}_{u}; \{IO_k\}_{k=1}^K \Big) \\
         s.t. \;\; \mathcal{D}( \ 
                        \hat{\textrm{F}}_{o}, \textrm{F}_{u} ) \ 
                        = 0  % \;\; \
        %  \hat{\textrm{F}}_{opt}^i(I_k^i) = \ 
        %         O_k^i , \\
            % \forall  \; k \;  \in 1...K 
    \end{split}
\end{equation}

where $\mathcal{D}(\cdot) \in \{ 0, 1 \}$ is an indicator function that measures if the model optimized program $\hat{F}_o$ is functionally equivalent to the original one $\p$. 
An intuitive choice of $\mathcal{D}$ is directly executing $\p$ and $\hat{F}_o$ on the test cases $\{ IO \}_{k=1}^K$ and checking the outputs equivalence. The cost function $\mathcal{C}(\cdot)$ considers metrics like runtime and the number of instructions in a program.
We will present more details in \cref{sec:setup}.
%, and could be computed by directly executing the optimized program on the test cases $\{ IO \}_{k=1}^K$. 
%Additionally, we explain how $\mathcal{D}$ may be calculated by using the suite of test cases, or by incorporting other methods like SMT-based solvers. 
For conciseness, we will abbreviate $\mathbfcal{C} (\hat{\textrm{F}}_{o}; \{IO_k\}_{k=1}^K)$ as $\mathbfcal{C} (\hat{\textrm{F}}_{o})$.

@@

%We perform a two-step process for learning program optimization.
In this section we elaborate on our approach that learns to optimize real-world programs. 
We first describe the underlying neural model before introducing the learning algorithm.

@@

Our program optimization model $f_{\theta}$ is a neural sequence-to-sequence network, where input (unoptimized) and output (optimized) programs are represented as sequences of tokens.
Specifically, $f_{\theta}$ is parameterized with a standard Transformer-based encoder \citep{vaswani2017attention}, but with a LSTM-based decoder \citep{hochreiter1997long,chen-etal-2018-best} instead.
This was done to speed-up prediction within the iterative training loop described in \S\ref{sec:hillclimbing}, as Transformers suffer from quadratic decoding time in sequence length \citep{zhang-etal-2018-accelerating}. We defer more implementation-specific details for our experiments to Section \ref{sec:experiments}.

@@

\subsection{Learning Algorithm}
\label{sec:hillclimbing}

@@

%We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
%It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
\cref{alg:hill_climbing} illustrates the hill-climbing algorithm for learning to optimize programs.
It consists of two steps, an exploration step (lines xxx-xxx) where the model seeks to discover alternative optimizations that are more efficient than the compiler generated targets used in pre-training, and a learning step (lines xxx-xxx), where the model parameters are updated using newly discovered optimized programs\pcyin{TODO: mark those two steps in the Algorithm}.
First, in the exploration step, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and execute $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
%Each output from the batch is executed on the test cases to test equivalence and its performance characteristics are measured and profiled. 
If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
% \pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.} \alex{not only 1 per input}
After the exploration step is taken, in the learning step, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

\paragraph{Discussion} \cref{alg:hill_climbing} bears some resemblance to self-imitation learning \cite{oh2018self} in which an agent is trained to imitate previously optimal actions in a reinforcement learning context. This is performed on individual actions in an episode; whereas, we train on the entire sequence. Learning with this algorithm may also be interpreted as imitation learning on an expert policy that iteratively improves. 
Finally, we remark that an alternative approach for learning program optimizations could be policy gradient methods.
In this paper we adopt a simpler hill-climbing approach as we observed in initial pilot experiments that the policy gradient was unstable, due to the general sparsity of rewards and the long trajectories of programs.

% (\pyin{TODO: xxx} tokens in average).
% Such instability of RL-based learning algorithms confirms with existing research on program synthesis \citep{bunel2018leveraging}.

% More formally, it could be interpreted as an extreme case of normalized off-policy reinforcement learning. We refer interested readers to Apendix xxx for more details \alex{TODO: write out and put in bibliography.}.

@@

% \textcolor{red}{=== Begin of Alex's Version ===}

% %\pcyin{When explaining a model/algorithm, it's important to first briefly describe the intuition behind each component/sub-section at the very beginning. Can we say something like the follows here: With the model pre-trained on compiler-optimized examples, we then aim to learn more efficient optimization strategies by directly minimizing the expected cost of model-optimized programs.}

% %%Commented by Pengcheng: 
% Policy gradient algorithms seem like a natural fit for learning how to optimize programs; however, methods based on maximum-likelihood estimation, or imitation learning, provide advantages in terms of stability. Training large neural networks with policy gradient algorithms can be difficult to tune. These difficulties are often attributed to the gradient's high variance which is exacerbated by other issues in program synthesis: a general sparsity of rewards, credit assignment over long sequences, and brittleness of program semantics.

% % \alex{I think this point here contradicts my earlier point about needing to sample from a grammar.... If a grammar really isn't that necessary after pre-training, then is brittle syntax really an issue?? It may be good to leave this out or accept either the nuance proposed / contradiction}
% % \pcyin{Perhaps just removing ``brittleness of program syntax'' is good?}

% %Commented by Pengcheng: 
% If an expert policy or expert examples are available, imitation learning can be desirable, because of its stability. Impressive results have been achieved in the game of computer Go and other games by minimizing the cross-entropy between a policy and an expert-like Monte Carlo tree search policy \cite{silver2017mastering}. In program synthesis from I/O example tasks, supervised learning on example programs has been shown to be competitive, and at times even superior to RL \cite{bunel2018leveraging}. The effectiveness of supervised, or imitation learning, on program synthesis tasks is even more surprising, given the programs used as targets were generated from random generation from a grammar. Given that real-world programs demonstrate a degree of naturalness, and thus, a higher degree of predictability and repetition \cite{hindle2012naturalness}, one may expect imitation learning to generally perform even better in practice on general program synthesis tasks. \alex{I feel these 2 sentences are good points but could be a little too much detail.}\pcyin{I agree. We could try shortening this paragraph once Section 6 is nearly done.}

% % The experiments performed in \cite{bunel2018leveraging} showed that in program synthesis tasks in the Karel domain, maximum-likelihood training on just one correct program example was competitive with reinforcement learning on a reward function defined by executed program outputs. Moreover, maximum-likelihood training outperformed reinforcement learning when the beam size for sampling candidates was large. The result may be surprising as maximum-likelihood training ignores issue of  \textit{program aliasing} in which multiple correct programs, not just one, is capable of satisfying the program specification, and may over-fit on the maximum-likelihood target.

% % Commented by Pengcheng

% Inspired by the success of imitation learning in program synthesis, in this paper we propose an intuitive hill-climbing algorithm, a special case of imitation learning, for learning program optimizations. 
% %We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
% %It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
% It consists of a two-step process. 
% First, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
% For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and run $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
% %Each output from the batch is executed on the test cases to test equivalence and its performance characteristics are measured and profiled. 
% If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
% \pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
% After the exploration step is taken, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

% Algorithm \ref{alg:hill_climbing} bears some resemblance to self-imitation learning \cite{oh2018self} \pcyin{can you briefly explain why they are similar?}; except we train on the entire sequence, instead of individual actions within a trajectory. Learning with this algorithm may be interpreted as imitation learning on an expert policy that iteratively improves. In our appendix, we also show how this is an extreme case of normalized off-policy reinforcement learning \alex{TODO: write out and put in bibliography.}.

% \textcolor{red}{=== End of Alex's Version ===}

\begin{algorithm}[t]
\begin{algorithmic}
\STATE Initialize model $f$ parameters $\theta$ from pre-trained model \\
\STATE Initialize dataset of program function pairs and test cases: \\  
    $D = D_o =  \Big\{ \big( \
    \{IO^i\}_{k=1}^K, \
    \textrm{F}_{u}^i, \
    \textrm{F}_{o}^i, \
    % \textit{LiveOut}^i\
    \big) \ 
    \Big\}_{i=1...N}$ \\
 \WHILE{not converged}
  \STATE Sample a batch $B_{\mathrm{ex}}$ from $D_o$\\
  \FOR{ $\big( \
    \{IO^i\}_{k=1}^K, \
    \textrm{F}_{u}^i, \
    \textrm{F}_{o}^i, \
    % \textit{LiveOut}^i\
    \big)$  in $B_{\mathrm{ex}}$} 
    \STATE  sample $\hat{\textrm{F}}_{o}^i \; \sim \; f_{\theta}(\textrm{F}_{u}^i)$ \\
    \STATE calculate 
    $\mathcal{C}(\textrm{F}_{o}^i)$, 
    and $\mathcal{D} (\hat{\textrm{F}}_{o}^i, \textrm{F}_{o}^i)$ 
    \IF{$\mathcal{D} (\hat{\textrm{F}}_{o}^i, \textrm{F}_{o}^i) == 0$ and \ 
    $\mathcal{C}(\hat{\textrm{F}}_{o}^i) < \ 
    \mathcal{C}(\textrm{F}_{o}^i) $ }
        \STATE  Replace $\textrm{F}_{o}^i$ with sample $\hat{\textrm{F}}_{o}^i$ in $D$ \\
    \ENDIF
  \ENDFOR
  \STATE Sample a batch $B_{\mathrm{tr}}$ from $D$\\
  \STATE Update $\theta$ via supervised learning on $B_{\mathrm{tr}}$ from $D$ \\
 \ENDWHILE
 \caption{Hill Climbing for Program Optimization}
 \label{alg:hill_climbing}
\end{algorithmic}
\end{algorithm}

@@

\paragraph{Actor-Learner Architecture}

% \pcyin{Maybe briefly explain why we use multiprocessing: because evaluating $\mathcal{D}(\cdot)$ is slow. Would be better to give a rough runtime number.}
To perform our experiments at scale, we utilize an actor-learner set up  \citep{liang2018memory, espeholt2018impala}. This was because the time it took to evaluate $\mathcal{C}$ and $\mathcal{D}$ were costly. In our actor-learner architecture, before the training process begins, multiple actor threads are created which inherit the parameters of the parent learner and the dataset. For every iteration, each of the actor threads independently samples $B_{\textrm{ex}}$ from the distribution of the inherited model.

@@

% Utilizing the actor-learner architecture was pivotal in being able to scale the speed of training. Otherwise, during our experiments, performing synchronized learning was slow \alex{is there a good way to quantify this ?}.

@@

% \pcyin{This could be commented out as we mention them as titles of sub-sections.}

\alex{@graham The 4 experimental questions are listed in the subsections and paragraphs after 6.1}

% We are primarily interested in answering the following research questions: 
% (1) Are neural sequence-to-sequence models capable of learning generalized semantics-preserving program optimizations ? We seek to measure this by investigating the model's ability to transduce correct programs that outperform unoptimized programs

% (2) Can neural sequence model pre-trained on demonstrations of an optimizing compiler like \texttt{GCC} (\cref{sec:sec:results_training_with_test_cases}) learn to outperform the optimization levels for which it was pre-trained ?

% (3) In learning to optimize real world programs, how much benefit do more involved methods of program verification, such as using SMT solvers, provide ?

% (4) What kind of optimization patterns does the neural optimizer learn (\cref{sec:sec:opt_examples})?

@@

\paragraph{Measuring Program Performance and Correctness} 
%\pcyin{We need to keep in mind that since we chose not to report results from training with SMT-solvers, we need to be careful to distinguish between the correctness function $\mathcal{D}$ used in \cref{alg:hill_climbing} for training and the SMT-bsed post-verification used in Table 2.}
In our experiments, we calculate the cost function $\mathcal{C}$ (E.q. \ref{eqn:optimizaiton_goal}) first by calculating the sum of all expected CPU clock-cycles for each opcode in the synthesized assembly. 
Then, we execute the entire test suite on the synthesized program and count the number of instructions actually executed. Our cost function is the sum of these values.%
\footnote{We chose to use these these methods of approximating runtime, as the actual measured runtime of the program re-writes may both be hardware-dependent and inconsistent between executions due to idiosyncrasies of the hardware used for experiments.}

% For each unoptimized program $F_u$, have a corresponding set of input examples ${\{IO_k\}}_{k=1...K}$ generated through random generation. \pcyin{The last sentence is a bit confusing?}

@@

\paragraph{Evaluation}
Given an evaluation set $E$ of unoptimized programs, we report the percentage of programs in $E$ whose model-optimized programs outperform the compiler-optimized results as measured by the cost function $\mathcal{C}(\cdot)$.
Specifically, for each unoptimized program, we generate model-optimized candidates using a beam-size of 10, executing all on the test suite $\{IO\}_{k}^K$ to evaluate the cost and semantic equivalence, and report results on the best-performing correct sample (if one was available). 
We use the SMT-based verifier ($\correctfuncsmt/$) to measure program equivalence. 
%We compare the model optimized program with two compiler configurations.

@@

% \subsection{Training Only with test cases}
%\subsection{Training using $\bm{\mathcal{D}_\textrm{I/O}}$}
%\label{sec:sec:results_training_with_test_cases}

@@

\subsection{Are  neural  sequence-to-sequence  models  capable  of learning and discovering optimizations ?}

\cref{tab:z3_eval_results} shows our main results.
We report the percentage of semantic-preserving optimizations that outperform the original program (\textbf{Beat Unoptimized}) and the optimizer (\textbf{Beat GCC}).
